Analiza vremenskih nizova zasnovana na kompleksnim mrežama

### 2017-06-06 tex/diplomski.tex
Hrvatski

% Ispis stranice s napomenom o umetanju izvornika rada. Uklonite naredbu \izvornik ako želite izbaciti tu stranicu.
\izvornik

@@

  Ova relacija specifična je po tome što je ($\forall x, y, z \in \Omega$):
  \begin{itemize}
%    \item \textit{irreflexive}: every entity $x$ is not preferrable over itself, %($\neg \left( x \succ x \right)$),
    \item \textit{irrefleksivna}: $\neg \left( x \succ x \right)$ --- za nijedno dobro ne vrijedi da je više preferirano od samog sebe,
%    \item \textit{asymmetrical}: if $x$ is preferrable over some $y$, then $y$ is not preferrable over $x$, %($x \succ y \Rightarrow \neg \left( y \succ x \right)$),
    \item \textit{asimetrična}: $x \succ y \Rightarrow \neg \left( y \succ x \right)$ --- ako je $x$ više preferirano od $y$, onda $y$ nije više preferirano od $x$,
%    \item \textit{transitive}: if $x$ is preferrable over $y$, and $y$ is preferrable over $z$, then $x$ is also preferrable over $z$, %($x \succ y \wedge y \succ z \Rightarrow x \succ z$),
    \item \textit{tranzitivna}: $x \succ y \wedge y \succ z \Rightarrow x \succ z$ --- ako je $x$ više preferirano od $y$, te $y$ više preferirano od $z$, tada je i $x$ više preferirano od $z$,
%    \item \textit{transitive in incomparability} (noting that $x$ and $y$ may be \textit{incomparable}, i.e. neither $x$ is preferrable over $y$, nor $y$ is preferrable over $x$): if $x$ is incomparable with $y$, and $y$ is incomparable with $z$, then $x$ is also incomparable with $z$.
    \item \textit{tranzitivna po indiferentnosti}: $x \sim y \wedge y \sim z \Rightarrow x \sim z$ --- ako je $x$ podjednako preferirano kao i $y$, te $y$ podjednako preferirano kao i $z$, tada je i $x$ podjednako preferirano kao i $z$.
  \end{itemize}

@@

%  Utility function $U\colon \Omega \to \mathbb{R}$ is mapping from entities to real numbers, in such way that order of the mapping corresponds to the preference order of the entities, i.e. $\forall x, y \in \Omega, U(x) > U(y) \Rightarrow x \succ y$.
%  One such mapping is obtained by using potential method that is described later in this paper.
%  In addition to ordering of the entities, utility function also provides the intensity of preference for a particular entity, hence it is more informative when it comes to the decision making.
  Funkcija korisnosti $U\colon \Omega \to \mathbb{R}$ je preslikavanje iz skupa dobara u skup realnih brojeva, na način da poredak preslikanih realnih brojeva odgovara poretku dobara prema individualnoj preferenciji, tj. vrijedi $\forall x, y \in \Omega, U(x) > U(y) \Leftrightarrow x \succ y$.
  Jedno takvo preslikavanje ostvaruje se korištenjem metode potencijala koja je opisana u poglavlju \ref{sec:metpot}.
  Uz to što određuje poredak dobara prema individualnoj preferenciji, funkcija korisnosti također unosi i mjeru intenziteta preferencije prema nekoj vrijednosnici, što znači da nosi više informacija od same relacije.

@@

  \section{Metoda potencijala}
  \label{sec:metpot}
  From previously obtained graph it is possible to tell which pair of assets has the highest preference flow.
  However, it is not yet possible to directly tell which are the most or least preferrable assets, or obtain the measure of preference for individual assets.
  To calculate preferences for each node in the graph, we use the potential method\cite{caklovic}.
  The potential of a node corresponds to difference in amount of flow going in and out of the node.

  %  A concise summary of the method is as follows:
  %  \begin{enumerate}
  %    \item 
  For the observed graph $\graph{G}$, let there be a total of $N$ nodes, and maximum of $E = \binom{N}{2}$ edges, in case of a complete graph.
  If $\graph{G}$ is not complete, we complete it by adding edges to it with weight 0 (direction doesn't matter),
  thus forming a complete graph $\graph{G}$ with $\binom{N}{2}$ edges.
  % In our case, graphs will be incomplete for most of the time, but they may be \textsl{supplemented} (TODO: opposite of `reduced') to complete graphs, if we treat missing edges as edges of weight 0 (direction doesn't matter).

  %    \item
  Let $\matr{B}$ be the $E \times N$ incidence matrix of $\graph{G}$.
  %    $\matr{B}$ has following properties:
  %    \begin{enumerate}
  %      \item each row corresponds to an edge in the graph, and each column to a node,
  %      \item for every edge in the graph going from node $i$ to node $j$, there is a corresponding row that has $-1$ and $1$ in columns that correspond to nodes $i$ and $j$ respectively,
  %      \item the remainder of elements in the matrix are zeros.
  %    \end{enumerate}
  %
  %    \item
  Let $\matr{f}$ be $E \times 1$ vector that contains edge weights (i.e. preference flows).
  Order of the edges must be the same as order of the edges in $\matr{B}$.
  As mentioned before, in place of missing edges we simply put 0.
  %    
  %    \item
  Let $\matr{\phi}$ be $N \times 1$ vector that contains potentials of each node, in order that is the same as order of the nodes in $\matr{B}$.

  %    \item 
  Now, if $\graph{G}$ was consistent, then $\matr{B}$, $\matr{\phi}$, and $\matr{f}$ would satisfy the equation
  \begin{equation}
  \label{eq:flowformula}
  \matr{B} \matr{\phi} = \matr{f}.
  \end{equation}
  Equation (\ref{eq:flowformula}) states that the difference between potential of any two nodes should result in weight of the edge between them.
  This is possible only for consistent graphs, and most of the time our graphs will be inconsistent.
  In that case we try to find an approximate solution $\matr{\phi^*}$ that minimizes the square error:
  \begin{gather}
  \matr{\phi^*} = \argmin_{\matr{\phi}} \left\{ \left \lVert \matr{B} \matr{\phi} - \matr{f}\textsl{} \right \rVert ^ 2 \right\} \nonumber \\ 
  \Downarrow \nonumber \\
  \label{eq:derivative}
  \frac{\partial \left \lVert \matr{B} \matr{\phi^*} - \matr{f} \right \rVert ^ 2}{\partial \matr{\phi^*}} = \mathbf{0}.
  \end{gather}
  Solving (\ref{eq:derivative}) via commonly used techniques of matrix calculus brings us to the following equation:
  \begin{gather}
  \label{eq:flow}
  2 \matr{B}^\T \left[\matr{B} \matr{\phi^*} - \matr{f} \right] = \mathbf{0} \nonumber \\
  \matr{B}^\T \matr{B} \matr{\phi^*} = \matr{B}^\T \matr{f}.
  \end{gather}
  Equation (\ref{eq:flow}) determines $\matr{\phi^*}$ up to a constant (i.e. solution has one degree of freedom), so the following constraint is also included:
  \begin{equation}
  \label{eq:sumiszero}
  \matr{j}^\T \matr{\phi^*} = 0
  \end{equation}
  where $\matr{j}$ is vector of ones with same dimension as $\matr{\phi^*}$.
  This ensures an unique solution for which total amounts of positive and negative potential will be equal.

  %    \item
  Joining the previous two equations together by adding (\ref{eq:sumiszero}) to each row in (\ref{eq:flow}) results in:
  \begin{align}
  \matr{B}^\T \matr{B} \matr{\phi^*} + \matr{J} \matr{\phi^*} &= \matr{B}^\T \matr{f} \nonumber \\
  \label{eq:joined}
  \left[\matr{B}^\T \matr{B} + \matr{J} \right] \matr{\phi^*} &= \matr{B}^\T \matr{f},
  \end{align}
  where $\matr{J}$ is a matrix of ones with same dimension as $\matr{B}^\T \matr{B}$.
  Finally, solving (\ref{eq:joined}) for $\matr{\phi^*}$ gives us:
  \begin{equation}
  \label{eq:final}
  \matr{\phi^*} = \left[\matr{B}^\T \matr{B} + \matr{J} \right]^{-1} \matr{B}^\T \matr{f}.
  \end{equation}
  %    \item 
  Furthermore, the term $\left[ \matr{B}^\T \matr{B} + \matr{J} \right]^{-1}$ in (\ref{eq:final}) can be simplified to $\frac{1}{N} \matr{I}$ due to $\matr{B}^\T \matr{B}$ being the Laplace matrix of a complete graph; thus, we can simplify (\ref{eq:final}) some more:
  \begin{equation}
  \matr{\phi^*} = \frac{1}{N} \matr{B}^\T \matr{f},
  \end{equation}
  to get as computationally optimal expression as possible.

  %    \item
  Afterwards, we can calculate the consistent reconstruction $\matr{f^*}$ of preference  flow by simply plugging back $\matr{\phi^*}$ into (\ref{eq:flowformula}):
  \begin{equation}
  \matr{f^*} = \matr{B} \matr{\phi^*}.
  \end{equation}
  The reconstructed preference flow $\matr{f^*}$ compared to the original preference flow $\matr{f}$ may even contain some new and/or lose some old edges.
  In addition, $\matr{B}$, $\matr{\phi^*}$, and $\matr{f^*}$ now describe a consistent graph $\graph{G}^*$.
  It is now possible to define a consistency measure $\kappa$ as follows:
  \begin{equation}
  \label{eq:consistency}
  \kappa = \frac{\left \lVert \matr{f^*} \right \rVert}{\left \lVert \matr{f} \right \rVert}.
  \end{equation}
  Equation (\ref{eq:consistency}) represents the cosine of the angle between $\matr{f}$ and $\matr{f^*}$ in the column space of matrix $\matr{B}$.
  Consistency measure $\kappa$ tells us how consistent graph $\graph{G}$ was, compared to the $\graph{G}^*$.
  $\kappa$ ranges from 0 to 1, with 0 meaning full inconsistency (virtually unreachable), and 1 meaning full consistency.

@@

  Let there be total of $N$ assets in $D$ days.
  Let price of asset $i$ at the time step $t$ be $a_i^{(t)}$, for $i \in {\left[1, 2, \ldots, N\right]}$ and $t \in {\left[0, 1, \ldots, D-1\right]}$.
  The log prices $b_i^{(t)}$, and log price differences $c_{i,j}^{(t)}$ between assets $i$ and $j$ are obtained as follows:
  \begin{equation} b_i^{(t)} = \log\left(a_i^{(t)}\right) \end{equation}
  \begin{equation} c_{i,j}^{(t)} = b_i^{(t)} - b_j^{(t)}, \end{equation}
  and rolling means $m_{i,j}^{(t)}$ and standard deviations $d_{i,j}^{(t)}$ of log price differences over the past time window of size $T$ are obtained as follows:

  \begin{equation}
  \label{eq:mean}
  m_{i,j}^{(t)} = \frac{1}{T}\sum_{\tau = t - T}^{t - 1} c_{i,j}^{(\tau)}
  \end{equation}
  \begin{equation}
  \label{eq:deviation}
  d_{i,j}^{(t)} = \sqrt{\frac{1}{T}\sum_{\tau=t - T}^{t - 1} \left(c_{i,j}^{(\tau)} - m_{i,j}^{(t)} \right)^2}.
  \end{equation}

  Note that in summation used in (\ref{eq:mean}, \ref{eq:deviation}) time step $t$ was intentionally excluded, therefore summation goes only to $t - 1$.
  We use these calculations as basis for creating the portfolio.

  Note that period over which means and standard deviations of log price differences are calculated does not include time step $t$.
  Also note that calculating them separately for each time step $t$ is rather computationally inefficient when dealing with rolling windows of data.
  Therefore, it is advisable to use a rolling algorithm as described in the appendix.
  On that note, $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ may be more efficiently stored if stored contiguously in memory as a matrix, using following coding scheme: a pair $(i, j)$, where $i < j$, should be encoded to $k$ as:
  \begin{equation} k = N \cdot (i - 1) + j - 1 - \left. i \cdot (i + 1) \middle/ 2\right., \end{equation}
  and decoded from $k$ as:
  \begin{equation} i = \left\lfloor N + 1/2 - \sqrt{(N + 1/2)^2 - 2(N + k)} \right\rfloor, \end{equation}
  \begin{equation} j = k + i \cdot \left.(i + 1) \middle/ 2\right. - N \cdot (i - 1) + 1. \end{equation}
  An example of proposed coding is shown on figure \ref{fig:coding}.

  \begin{figure}[h]
    \centering
    \begin{minipage}{0.5\textwidth}
      \centering
      \begin{tabular}{C|CCCCC}
        i/j & 1 & 2 & 3 & 4 & 5 \\ \hline
        1 & \cdot & 0 & 1 & 2 & 3 \\
        2 & \cdot & \cdot & 4 & 5 & 6 \\
        3 & \cdot & \cdot & \cdot & 7 & 8 \\
        4 & \cdot & \cdot & \cdot & \cdot & 9 \\
        5 & \cdot & \cdot & \cdot & \cdot & \cdot
      \end{tabular}
    \end{minipage}%
    \begin{minipage}{0.5\textwidth}
      \centering
      \begin{tabular}{C|CCCCCCCCCC}
       i & 1 & 1 & 1 & 1 & 2 & 2 & 2 & 3 & 3 & 4 \\
       j & 2 & 3 & 4 & 5 & 3 & 4 & 5 & 4 & 5 & 5 \\ \hline
       k & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 \\
%      k & i & j \\ \hline
%      0 & 1 & 2 \\
%      1 & 1 & 3 \\
%      2 & 1 & 4 \\
%      3 & 1 & 5 \\
%      4 & 2 & 3 \\
%      5 & 2 & 4 \\
%      6 & 2 & 5 \\
%      7 & 3 & 4 \\
%      8 & 3 & 5 \\
%      9 & 4 & 5 \\
      \end{tabular}
    \end{minipage}
    \caption{Example of the proposed coding scheme, for $N = 5$. A dot $(\cdot)$ indicates that that combination is not used.}
    \label{fig:coding}
  \end{figure}

  \section{Konstrukcija grafa toka preferencija}
  \label{sub:creating-graph}
  Using the obtained $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ it is now possible to create a graph of preference flow among assets for each time step $t$.
  Considering one time step $t$, we find all such pairs of assets $(i,j)$ that satisfy:
  \begin{equation}
  \label{eq:thresh}
  \left| c_{i,j}^{(t)} - m_{i,j}^{(t)} \right| > \alpha \cdot d_{i,j}^{(t)},
  \end{equation}
  i.e. current log price difference is at least $\alpha$ deviations distant from mean value of the past time window.
  An illustration is shown on Fig. \ref{fig:devmag}.
  %  Parameter $\alpha$ determines how many pairs of assets should constitute the graph at current time step.
  Afterwards, for each observed pair $(i,j)$ that exceeds the threshold we add into graph vertices $i$ and $j$, with a weighed edge of weight $w_{i,j}^{(t)}$ going from $i$ to $j$.
  Weight $w_{i,j}^{(t)}$ is obtained as:
  \begin{equation}
  \label{eq:weight}
  w_{i,j}^{(t)} = \left. \left(c_{i,j}^{(t)} - m_{i,j}^{(t)}\right) \middle/ d_{i,j}^{(t)} \right..
  \end{equation}

>>>>

% Ispis stranice s napomenom o umetanju izvornika rada. Uklonite naredbu \izvornik ako želite izbaciti tu stranicu.
%\izvornik

@@

  Ova relacija specifična je po tome što je:
  \begin{itemize}
%    \item \textit{irreflexive}: every entity $x$ is not preferrable over itself, %($\neg \left( x \succ x \right)$),
    \item \textit{irrefleksivna}: $\forall x \in \Omega\colon \neg \left( x \succ x \right)$ --- za nijedno dobro ne vrijedi da je više preferirano od samog sebe,
%    \item \textit{asymmetrical}: if $x$ is preferrable over some $y$, then $y$ is not preferrable over $x$, %($x \succ y \Rightarrow \neg \left( y \succ x \right)$),
    \item \textit{asimetrična}: $\forall x, y \in \Omega\colon x \succ y \Rightarrow \neg \left( y \succ x \right)$ --- ako je $x$ više preferirano od $y$, onda $y$ nije više preferirano od $x$,
%    \item \textit{transitive}: if $x$ is preferrable over $y$, and $y$ is preferrable over $z$, then $x$ is also preferrable over $z$, %($x \succ y \wedge y \succ z \Rightarrow x \succ z$),
    \item \textit{tranzitivna}: $\forall x, y, z \in \Omega\colon x \succ y \wedge y \succ z \Rightarrow x \succ z$ --- ako je $x$ više preferirano od $y$, te $y$ više preferirano od $z$, tada je i $x$ više preferirano od $z$,
%    \item \textit{transitive in incomparability} (noting that $x$ and $y$ may be \textit{incomparable}, i.e. neither $x$ is preferrable over $y$, nor $y$ is preferrable over $x$): if $x$ is incomparable with $y$, and $y$ is incomparable with $z$, then $x$ is also incomparable with $z$.
    \item \textit{tranzitivna po indiferentnosti}: $\forall x, y, z \in \Omega\colon x \sim y \wedge y \sim z \Rightarrow x \sim z$ --- ako je $x$ podjednako preferirano kao i $y$, te $y$ podjednako preferirano kao i $z$, tada je i $x$ podjednako preferirano kao i $z$.
  \end{itemize}

@@

%  Utility function $U\colon \Omega \to \mathbb{R}$ is mapping from entities to real numbers, in such way that order of the mapping corresponds to the preference order of the entities, i.e. $\forall x, y \in \Omega, U(x) > U(y) \Rightarrow x \succ y$.
%  One such mapping is obtained by using potential method that is described later in this paper.
%  In addition to ordering of the entities, utility function also provides the intensity of preference for a particular entity, hence it is more informative when it comes to the decision making.
  Funkcija korisnosti \engl{utility function} $U\colon \Omega \to \mathbb{R}$ je preslikavanje iz skupa dobara u skup realnih brojeva, na način da poredak preslikanih realnih brojeva odgovara poretku dobara prema individualnoj preferenciji, tj. vrijedi $\forall x, y \in \Omega \colon U(x) > U(y) \Leftrightarrow x \succ y$.
  Jedno takvo preslikavanje ostvaruje se korištenjem metode potencijala koja je opisana u poglavlju \ref{sec:metpot}.
  Uz to što određuje poredak dobara prema individualnoj preferenciji, funkcija korisnosti također unosi i mjeru intenziteta preferencije prema nekoj vrijednosnici, što znači da nosi više informacija od same relacije.

@@

  \section{Metoda potencijala}
  \label{sec:metpot}
%  From previously obtained graph it is possible to tell which pair of assets has the highest preference flow.
%  However, it is not yet possible to directly tell which are the most or least preferrable assets, or obtain the measure of preference for individual assets.
%  To calculate preferences for each node in the graph, we use the potential method\cite{caklovic}.
%  The potential of a node corresponds to difference in amount of flow going in and out of the node.
  Iz prethodno dobivenog grafa moguće je utvrditi u kojem paru vrijednosnica je tok preferencije najveći.
  Ipak, još uvijek nije moguće izravno utvrditi koja konkretna vrijednosnica je najviše ili najmanje poželjna, ili odrediti individualne mjere preferencije za svaku vrijednosnicu.
  Kako bi to bilo moguće korištena je metoda potencijala koja je opisana u nastavku.

  %  A concise summary of the method is as follows:
  %  \begin{enumerate}
  %    \item 
%  For the observed graph $\graph{G}$, let there be a total of $N$ nodes, and maximum of $E = \binom{N}{2}$ edges, in case of a complete graph.
%  If $\graph{G}$ is not complete, we complete it by adding edges to it with weight 0 (direction doesn't matter),
%  thus forming a complete graph $\graph{G}$ with $\binom{N}{2}$ edges.
  % In our case, graphs will be incomplete for most of the time, but they may be \textsl{supplemented} (TODO: opposite of `reduced') to complete graphs, if we treat missing edges as edges of weight 0 (direction doesn't matter).

  \subsection{Računanje potencijala čvorova}

  Neka je za promatrani graf $\graph{G}$ ukupno $N$ čvorova, te najviše $E = \binom{N}{2}$ bridova, ukoliko se radi o potpunom grafu.
  Ako $\graph{G}$ nije potpun graf, on se može dopuniti na način da se dodaju bridovi između čvorova koji nisu povezani, čija je težina jednaka 0, a sam smjer nije bitan.
  U nastavku se podrazumijeva da je $\graph{G}$ potpun graf.

  %    \item
%  Let $\matr{B}$ be the $E \times N$ incidence matrix of $\graph{G}$.
  Neka je $\matr{B}$ matrica incidencije grafa $\graph{G}$.
  Matrica incidencije je matrica za koju vrijedi:
  \begin{enumerate}
    \item broj redaka jednak je broju bridova $E$, i broj stupaca jednak je broju čvorova $N$;
    \item za svaki od $E$ bridova grafa postoji odgovarajući redak u matrici koji ga opisuje: ako brid ide od čvora $i$ do čvora $j$, tada taj redak sadrži $-1$ i $1$ u stupcima koji odgovaraju čvorovima $i$ i $j$;
    \item ostali elementi matrice su jednaki 0.
  \end{enumerate}
  %    $\matr{B}$ has following properties:
  %    \begin{enumerate}
  %      \item each row corresponds to an edge in the graph, and each column to a node,
  %      \item for every edge in the graph going from node $i$ to node $j$, there is a corresponding row that has $-1$ and $1$ in columns that correspond to nodes $i$ and $j$ respectively,
  %      \item the remainder of elements in the matrix are zeros.
  %    \end{enumerate}
  %
  %    \item
%  Let $\matr{f}$ be $E \times 1$ vector that contains edge weights (i.e. preference flows).
%  Order of the edges must be the same as order of the edges in $\matr{B}$.
%  As mentioned before, in place of missing edges we simply put 0.
  Neka je $\matr{f}$ vektor dimenzija $E \times 1$ koji sadrži težine bridova, tj. tokove preferencije,
  %    
  %    \item
  i neka je $\matr{\phi}$ vektor dimenzija $N \times 1$ koji sadrži potencijale čvorova, koji se žele pronaći.
  Redoslijed pripadnih čvorova i bridova moraju odgovarati redoslijedu čvorova i bridova u matrici incidencija $\matr{B}$.
%  Let $\matr{\phi}$ be $N \times 1$ vector that contains potentials of each node, in order that is the same as order of the nodes in $\matr{B}$.

  %    \item 
%  Now, if $\graph{G}$ was consistent, then $\matr{B}$, $\matr{\phi}$, and $\matr{f}$ would satisfy the equation
  U idealnom slučaju, kada je $\graph{G}$ konzistentan, $\matr{B}$, $\matr{\phi}$, i $\matr{f}$ zadovoljavaju jednadžbu:
  \begin{equation}
  \label{eq:flowformula}
  \matr{B} \matr{\phi} = \matr{f}.
  \end{equation}
%  Equation (\ref{eq:flowformula}) states that the difference between potential of any two nodes should result in weight of the edge between them.
%  This is possible only for consistent graphs, and most of the time our graphs will be inconsistent.
%  In that case we try to find an approximate solution $\matr{\phi^*}$ that minimizes the square error:
  Jednadžba (\ref{eq:flowformula}) tvrdi da razlike potencijala dvaju čvorova odgovaraju toku preferencije, odnosno težini brida koji povezuje te čvorove (do na predznak).
  Ovo je zadovoljivo samo za konzistentne grafove, što često i nije slučaj u ovom zadatku.
  U slučaju kada je graf $\graph{G}$ nekonzistentan, od interesa je pronaći rješenje $\matr{\phi^*}$ koje ima minimalnu kvadratnu pogrešku.
  Tako originalni problem postaje:
  \begin{gather}
  \label{eq:derivative}
  \matr{\phi^*} = \argmin_{\matr{\phi}} \left\{ \left \lVert \matr{B} \matr{\phi} - \matr{f}\textsl{} \right \rVert ^ 2 \right\} \quad
  \Leftrightarrow \quad 
  \frac{\partial \left \lVert \matr{B} \matr{\phi^*} - \matr{f} \right \rVert ^ 2}{\partial \matr{\phi^*}} = \mathbf{0}.
  \end{gather}
  Matričnim deriviranjem (\ref{eq:derivative}) dobiva se sljedeća jednadžba:
  \begin{gather}
  2 \matr{B}^\T \left[\matr{B} \matr{\phi^*} - \matr{f} \right] = \mathbf{0} \nonumber \\
  \label{eq:flow}
  \matr{B}^\T \matr{B} \matr{\phi^*} = \matr{B}^\T \matr{f}.
  \end{gather}
%  Equation (\ref{eq:flow}) determines $\matr{\phi^*}$ up to a constant (i.e. solution has one degree of freedom), so the following constraint is also included:
  Jednadžba (\ref{eq:flow}) određuje $\matr{\phi^*}$ do na konstantu, tj. rješenje ima jedan stupanj slobode.
  Stoga se dodaje sljedeće ograničenje:
  \begin{equation}
  \label{eq:sumiszero}
  \matr{j}^\T \matr{\phi^*} = 0
  \end{equation}
  gdje je $\matr{j}$ vektor jedinica istih dimenzija kao i $\matr{\phi^*}$.
  Jednadžba (\ref{eq:sumiszero}) osigurava jedinstveno rješenje u kojem je ukupan zbroj potencijala svih čvorova jednak nuli.

  Združivanjem prethodnih dviju jednadžbi u jednu, tako da se (\ref{eq:sumiszero}) doda svakom retku u (\ref{eq:flow}), dobiva se:
  \begin{align}
  \matr{B}^\T \matr{B} \matr{\phi^*} + \matr{J} \matr{\phi^*} &= \matr{B}^\T \matr{f} \nonumber \\
  \label{eq:joined}
  \left[\matr{B}^\T \matr{B} + \matr{J} \right] \matr{\phi^*} &= \matr{B}^\T \matr{f},
  \end{align}
  gdje je $\matr{J}$ matrica jedinica istih dimenzija kao i $\matr{B}^\T \matr{B}$.
  Konačno, rješavanjem (\ref{eq:joined}) po $\matr{\phi^*}$ dobiva se:
  \begin{equation}
  \label{eq:final}
  \matr{\phi^*} = \left[\matr{B}^\T \matr{B} + \matr{J} \right]^{-1} \matr{B}^\T \matr{f}.
  \end{equation}
  %    \item 
  Dodatno, izraz (\ref{eq:final}) sadrži u sebi Laplaceovu matricu $\matr{L} = \matr{B}^\T\matr{B}$.
  Za Laplaceovu matricu vrijedi:
  \begin{itemize}
    \item $(\matr{L})_{ii}$ je jednak broju susjeda čvora $i$,
    \item ostali elementi jednaki su $-1$.
  \end{itemize}
  Kako je graf $\graph{G}$ potpun, tj. broj susjeda svakog čvora jednak je $N - 1$, izraz $\left[ \matr{B}^\T \matr{B} + \matr{J} \right]^{-1}$ se može pojednostavniti u $\frac{1}{N} \matr{I}$; i stoga, se (\ref{eq:final}) može svesti na:
  \begin{equation}
  \label{eq:final2}
  \matr{\phi^*} = \frac{1}{N} \matr{B}^\T \matr{f},
  \end{equation}
  tako da se dobije računalno optimalan izraz.
  Jednadžba (\ref{eq:final2}) tvrdi da se rješenje u smislu najmanjih kvadrata jednadžbe (\ref{eq:flowformula}) dobije na način da se za svaki čvor pribroje težine svih bridova koji izlaze iz njega, i oduzmu težine svih bridova koji ulaze u njega.

  \subsection{Konzistentna rekonstrukcija grafa}

  Sada se na temelju dobivenog potencijala čvorova $\matr{\phi^*}$ može izračunati tok preferencije $\matr{f^*}$ koji je konzistentan s potencijalnom, tako da se u (\ref{eq:flowformula}) supstituira $\matr{\phi^*}$ umjesto $\matr{\phi}$:
  %    \item
%  Afterwards, we can calculate the consistent reconstruction $\matr{f^*}$ of preference flow by simply plugging back $\matr{\phi^*}$ into (\ref{eq:flowformula}):
  \begin{equation}
  \matr{f^*} = \matr{B} \matr{\phi^*}.
  \end{equation}
%  The reconstructed preference flow $\matr{f^*}$ compared to the original preference flow $\matr{f}$ may even contain some new and/or lose some old edges.
%  In addition, $\matr{B}$, $\matr{\phi^*}$, and $\matr{f^*}$ now describe a consistent graph $\graph{G}^*$.
%  It is now possible to define a consistency measure $\kappa$ as follows:
  Rekonsutrirani tok preferencije $\matr{f^*}$ u odnosu na originalni $\matr{f}$ može sadržavati neke nove, pa čak i izgubiti neke stare težine.
  Dodatno, $\matr{B}$, $\matr{\phi^*}$, i $\matr{f}$ opisuju konzistentan graf $\graph{G}^*$.
  Mjera konzistentnosti $\kappa$ definira se kao:
  \begin{equation}
  \label{eq:consistency}
  \kappa = \frac{\left \lVert \matr{f^*} \right \rVert}{\left \lVert \matr{f} \right \rVert}.
  \end{equation}
%  Equation (\ref{eq:consistency}) represents the cosine of the angle between $\matr{f}$ and $\matr{f^*}$ in the column space of matrix $\matr{B}$.
%  Consistency measure $\kappa$ tells us how consistent graph $\graph{G}$ was, compared to the $\graph{G}^*$.
%  $\kappa$ ranges from 0 to 1, with 0 meaning full inconsistency (virtually unreachable), and 1 meaning full consistency.
  Jednadžba (\ref{eq:consistency}) predstavlja kosinus kuta između $\matr{f}$ i $\matr{f^*}$ u prostoru određenom vektor stupacima matrice incidencije $\matr{B}$.
  Mjera konzistentnosti $\kappa$ opisuje koliko je sličan graf $\graph{G}$ s grafom $\graph{G}^*$, te je u rasponu od 0 do 1, gdje 0 znači potpunu nekonzistentnost (praktički nedostižno), a 1 potpunu konzistentnost.

@@

%  Let there be total of $N$ assets in $D$ days.
%  Let price of asset $i$ at the time step $t$ be $a_i^{(t)}$, for $i \in {\left[1, 2, \ldots, N\right]}$ and $t \in {\left[0, 1, \ldots, D-1\right]}$.
%  The log prices $b_i^{(t)}$, and log price differences $c_{i,j}^{(t)}$ between assets $i$ and $j$ are obtained as follows:
%  \begin{equation} b_i^{(t)} = \log\left(a_i^{(t)}\right) \end{equation}
%  \begin{equation} c_{i,j}^{(t)} = b_i^{(t)} - b_j^{(t)}, \end{equation}
%  and rolling means $m_{i,j}^{(t)}$ and standard deviations $d_{i,j}^{(t)}$ of log price differences over the past time window of size $T$ are obtained as follows:
  Neka u zadanom skupu podataka ima ukupno $N$ vrijednosnica kroz $D$ vremenskih koraka.
  Neka je cijena vrijednosnice $i$ u vremenskom koraku $t$ jednaka $a_i^{(t)}$, za $i \in {\left[1, 2, \ldots, N\right]}$ i $t \in {\left[0, 1, \ldots, D-1\right]}$.
  Log cijene $b_i^{(t)}$, i razlike log cijena po parovima $c_{i,j}^{(t)}$ dobivaju se na sljedeći način:
  \begin{equation} b_i^{(t)} = \log\left(a_i^{(t)}\right) \end{equation}
  \begin{equation} c_{i,j}^{(t)} = b_i^{(t)} - b_j^{(t)}, \end{equation}
  i srednja vrijednost $m_{i,j}^{(t)}$ i standardna devijacija $d_{i,j}^{(t)}$ razlika log cijena tijekom proteklog perioda duljine $T$ dobivene su kako slijedi:
  \begin{equation}
  \label{eq:mean}
  m_{i,j}^{(t)} = \frac{1}{T}\sum_{\tau = t - T}^{t - 1} c_{i,j}^{(\tau)}
  \end{equation}
  \begin{equation}
  \label{eq:deviation}
  d_{i,j}^{(t)} = \sqrt{\frac{1}{T}\sum_{\tau=t - T}^{t - 1} \left(c_{i,j}^{(\tau)} - m_{i,j}^{(t)} \right)^2}.
  \end{equation}

  Valja primijetiti kako je u izrazima sumiranja u (\ref{eq:mean}) i (\ref{eq:deviation}) vremenski korak $t$ namjerno izostavljen jer se radi nad proteklim periodom trajanja $T$, stoga sumiranje ide do $t - 1$.
%  Note that in summation used in (\ref{eq:mean}, \ref{eq:deviation}) time step $t$ was intentionally excluded, therefore summation goes only to $t - 1$.
  Ovi izračuni su temelj za daljnji tijek algoritma.
%  We use these calculations as basis for creating the portfolio.

%  Also note that calculating them separately for each time step $t$ is rather computationally inefficient when dealing with rolling windows of data.
%  Therefore, it is advisable to use a rolling algorithm as described in the appendix.
%  On that note, $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, and $d_{i,j}^{(t)}$ may be more efficiently stored if stored contiguously in memory as a matrix, using following coding scheme: a pair $(i, j)$, where $i < j$, should be encoded to $k$ as:
  Također valja primijetiti da bi izračunavanje srednjih vrijednosti i standardnih devijacija zasebno za svaki vremenski korak bilo računski jako neefikasno, pogotovo kad se radi o velikom skupu podataka.
  Stoga je preporučljivo koristiti algoritam s pomičnim prozorima za računanje, koji je opisan u dodatku (TODO).
  Što se tiče memorijske potrošnje, $c_{i,j}^{(t)}$, $m_{i,j}^{(t)}$, i $d_{i,j}^{(t)}$ su idejno tenzori dimenzija $(D - T) \times N \times N$.
  Međutim, sva tri tenzora su antisimetrična, u smislu da vrijedi $c_{i,j}^{(t)} = -c_{j,i}^{(t)}$, $m_{i,j}^{(t)} = -m_{j,i}^{(t)}$, i $d_{i,j}^{(t)} = -d_{j,i}^{(t)}$, stoga je polovina informacije sadržane u njima redundantna.
  Imajući to u vidu, ti tenzori se mogu `ispeglati' do dimenzije $(D - T) \times \left. N\cdot(N - 1) \middle/ 2\right.$, koristeći sljedeće preslikavanje kod kodiranja --- za kodiranje para $(i,j)$ u kod $k$ koristi se izraz:
  \begin{equation} k = N \cdot (i - 1) + j - 1 - \left. i \cdot (i + 1) \middle/ 2\right., \end{equation}
  a za dobivanje $(i, j)$ iz $k$ izrazi:
  \begin{equation} i = \left\lfloor N + 1/2 - \sqrt{(N + 1/2)^2 - 2(N + k)} \right\rfloor, \end{equation}
  \begin{equation} j = k + i \cdot \left.(i + 1) \middle/ 2\right. - N \cdot (i - 1) + 1. \end{equation}
%  An example of proposed coding for $N = 5$ is shown on figure \ref{fig:coding}.
  Na slici \ref{fig:coding} prikazano je kodiranje u slučaju $N = 5$ vrijednosnica.
